Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-04-27 overleaf/paper_SIMD.tex
Some more edits on Secs. 1-3. Still holding lock.

\begin{enumerate}
\item All statements in the program execute using the same protocol (sharing). That is, there is no share conversion. 
%\ishaq{This is not an assumption, this is our setting. We work in the single protocol world.}
\item There are two tiers of MPC instructions, local and remote. A local instruction (essentially just ADD) has cost $\beta$ 
and a remote instruction (e.g., MUX, MUL, SHL, etc.) has cost $\alpha$, where $\alpha >> \beta$. We assume that all remote 
instructions have the same cost.
%\ishaq{Don't we need to distinguish between local (cheap) vs. interactive (expensive) instructions? I don't see how we could assume this. -- Past suggestion from Vassilis: Ideally we should assume cost to be a convex function.}\ana{After discussion 1/12: replace single costs 1 with two levels of costs: $\alpha$ for non-local operations, e.g., MUL, and $\beta$ for local ops, e.g., ADD.}
\item  In MPC frameworks, executing $n$ operations ``at once'' in a single SIMD operation costs a lot less than executing those $n$ operations one by one.
Following Amdahl's law~\ana{citation!}, we write $\alpha = \frac{1}{s}p\alpha + (1-p)\alpha$, where $p$ is the fraction of execution time that benefits from amortization and $(1-p)$ 
is the fraction that does not, and $s$ is the available resource. Thus, $n\alpha = \frac{n}{s}p\alpha + n(1-p)\alpha$. 
For the purpose of the model we assume that $s$ is large enough and the term $\frac{n}{s}p\alpha$ amounts to a \emph{fixed cost} incurred regardless of
whether $n$ is $10,000$ or just $1$. (This models the cost of sending a packet from party A to party B for example.) The bottom line is that amortized execution 
of $n$ operations costs $\alpha_\mathit{fix} + n\alpha_{var}$ in contrast to unamortized execution, which costs $n\alpha_\mathit{fix} + n\alpha_{var}$ and, of course,
$\alpha_\mathit{fix} << n\alpha_\mathit{fix}$.
%We assume infinite parallel capacity---i.e., a single MPC-instruction costs as much as $N$ amortized instructions, namely $\alpha$ or $\beta$. 
%This is a standard assumption in Cryptographic Parallel RAM. ABY presents empirical support for this assumption~\ana{Add citations. PRAM, ABY.}
%\ishaq{Comment from Vassilis: We assume there are infinite parallel capacities, this is the standard assumption in PRAM (Cryptographic Parallel RAM).}\ana{PRAM assumption stays. We have to replace the unit cost of 1 with a suitable amortization function $f(n)$, which I don't think changes much in the cost modeling and analysis.}
\item MPC instructions scheduled in parallel benefit from amortization \emph{only if} they are the same instruction. Given our previous assumption,
2 MUL instructions can be amortized in a single SIMD instruction that costs $\alpha_\mathit{fix} + 2\alpha_\mathit{var}$, however a MUL and a MUX instruction 
still cost $2\alpha_\mathit{fix} + 2\alpha_\mathit{var}$ even when scheduled ``in parallel''. 
%MPC instructions scheduled in parallel benefit from amortization \emph{only if} they are the same instruction. Given our previous assumption,
%2 MUL instructions scheduled in parallel benefit from amortization and cost $\alpha$, however a MUL and a MUX instructions scheduled
%in parallel still cost $2\alpha$. 
%\ishaq{We need to reword this assumption to something like "$K$ parallel MUL costs much less than $K$ (because they get amortized), but any mix of $K$  MUL and MUX still cost roughly $K$. Specifically, we should take away the low constants (2 and 1) because we know for low constants this is not true.} \ana{Again, core assumption that MUL and MUX don't benefit from amortization stays. We have to replace constant costs with functions, as in (3).}
\end{enumerate}

@@

@@

\begin{figure*}
  \includegraphics[width=1.0\linewidth]{figs_paper_SIMD/compiler_framework.png}
  \caption{Compiler Framework.} 
  \label{fig:compiler_framework}
\end{figure*}

@@

\figref{fig:compiler_framework} presents an overview of our compiler. In this section, we describe several of the phases of the compiler. 
Sections~\secref{sec:vectorization} and~\secref{sec:div_and_conq} describe vectorization and divide-and-conquer.
We write $i,j,k$ to denote the loop nest: $i$ is the outermost loop, $j$, is immediately nested in $i$, and so on until $k$
and we use $I,J,K$ to denote the corresponding upper bounds. For simplicity, we write $A[i,j,k]$ to denote canonical access 
to an array element. In the program, canonical access is achieved via the standard row-major order formula: $(J*K)*i + K*j + k$.
To simplify the presentation we describe our algorithms in terms of three-element tuples $i,j,k$.
All discussion generalizes to arbitrarily large loop nests.
%\ana{Need to state this more precisely.}

\subsection{Syntax}

Our compiler imposes the following restrictions. We note that in some cases, the restrictions 
can be easily lifted and we plan to do so in future iterations of our compiler.

\begin{enumerate}
\item The program contains arbitrarily nested loops, however, loop bounds are fixed: \texttt{0 <= i < N}.
A standard restriction in MPC is that the bounds must be known at circuit-generation time.
\item Arrays are one-dimensional. N-dimensional arrays are linearized and accessed 
in row-major order and at this point the programmer is responsible for linearization
and access. 
\item Array subscrpts are plaintext values.
\item Our compiler allows for output (write) arrays, however it restricts
write access to \emph{canonical writes} along the dimensions of the array. I.e., \texttt{A[i,j] = ...}
where \texttt{i} and \texttt{j} loop over the two dimensions of \texttt{A} is allowed, 
but \texttt{A[i,j+2] = ...} is not allowed. Read access is arbitrary. 
\end{enumerate}

@@

\begin{enumerate}
  \item
        \textbf{Parsing}:
        Use Python's \texttt{ast} module to parse the input source code to a Python AST.
  \item
        \textbf{Syntax checking}:
        Ensure that the AST matches a restricted subset that our compiler supports.
        This step outputs an instance of the \texttt{restricted\_ast.Function} class, which represents our restricted subset of the Python AST.
  \item
        \textbf{3-address CFG conversion}:
        \benjamin{TODO: Is this a good amount of detail?}

@@

Then, the control-flow graph is \emph{linearized} into MPC source,
which has loops but no if-statements.
This means that both branches of all if-statements are executed,
and the MUX nodes determine whether to use results from the then-block or from the else-block.
The compiler linearizes the control-flow graph with a variation of breadth-first search.
Blocks with the ``merge condition'' property are only considered
the second time they are visited,
since that will be after both branches of the if-statement are visited.
Each time the compiler visits a block,
it adds the block's statements to the MPC source.
If the block ends in a for-instruction,
the compiler recursively converts the body and code after the loop to MPC source
and adds the for-loop and code after the loop to the main MPC source.
If the block does not end in a for-instruction,
the compiler recursively converts all successor branches to MPC source and
appends these to the main MPC source.

@@

An important component of our algorithm is the "lifting" of scalars to the corresponding loop dimensionality. 
For example, \texttt{d = S[i * D + j] - C[j]} equiv. to \texttt{d = S[i,j] - C[j]}, which gave rise to \texttt{N*D} subtraction operations in the sequential schedule, 
is lifted. The argument arrays \texttt{S} and \texttt{C} are lifted and the scaler \texttt{d} is lifted: \texttt{d[i,j] = S[i,j] - C[i,j]}.
The algorithm then detects that the statement can be vectorized.

>>>>

\begin{enumerate}
\item All statements in the program execute using the same protocol (sharing). That is, there is no share conversion. 
%\ishaq{This is not an assumption, this is our setting. We work in the single protocol world.}
\item There are two tiers of MPC instructions, local and remote. A local instruction (essentially just ADD) has cost $\beta$ 
and a remote instruction (e.g., MUX, MUL, SHL, etc.) has cost $\alpha$, where $\alpha >> \beta$. We assume that all remote 
instructions have the same cost.
%\ishaq{Don't we need to distinguish between local (cheap) vs. interactive (expensive) instructions? I don't see how we could assume this. -- Past suggestion from Vassilis: Ideally we should assume cost to be a convex function.}\ana{After discussion 1/12: replace single costs 1 with two levels of costs: $\alpha$ for non-local operations, e.g., MUL, and $\beta$ for local ops, e.g., ADD.}
\item  In MPC frameworks, executing $n$ operations ``at once'' in a single SIMD operation costs a lot less than executing those $n$ operations one by one.
Following Amdahl's law~\ana{citation!}, we write $\alpha = \frac{1}{s}p\alpha + (1-p)\alpha$, where $p$ is the fraction of execution time that benefits from amortization and $(1-p)$ 
is the fraction that does not, and $s$ is the available resource. Thus, $n\alpha = \frac{n}{s}p\alpha + n(1-p)\alpha$. 
For the purpose of the model we assume that $s$ is large enough and the term $\frac{n}{s}p\alpha$ amounts to a \emph{fixed cost} incurred regardless of
whether $n$ is $10,000$ or just $1$. (This models the cost of sending a packet from party A to party B for example.) The bottom line is that amortized execution 
of $n$ operations is $f(n) = \alpha_\mathit{fix} + n\alpha_{var}$ in contrast to unamortized execution $g(n) = n\alpha_\mathit{fix} + n\alpha_{var}$ and, of course,
$\alpha_\mathit{fix} << n\alpha_\mathit{fix}$.
%We assume infinite parallel capacity---i.e., a single MPC-instruction costs as much as $N$ amortized instructions, namely $\alpha$ or $\beta$. 
%This is a standard assumption in Cryptographic Parallel RAM. ABY presents empirical support for this assumption~\ana{Add citations. PRAM, ABY.}
%\ishaq{Comment from Vassilis: We assume there are infinite parallel capacities, this is the standard assumption in PRAM (Cryptographic Parallel RAM).}\ana{PRAM assumption stays. We have to replace the unit cost of 1 with a suitable amortization function $f(n)$, which I don't think changes much in the cost modeling and analysis.}
\item MPC instructions scheduled in parallel benefit from amortization \emph{only if} they are the same instruction. Given our previous assumption,
2 MUL instructions can be amortized in a single SIMD instruction that costs $\alpha_\mathit{fix} + 2\alpha_\mathit{var}$, however a MUL and a MUX instruction 
still cost $2\alpha_\mathit{fix} + 2\alpha_\mathit{var}$ even when scheduled ``in parallel''. 
%MPC instructions scheduled in parallel benefit from amortization \emph{only if} they are the same instruction. Given our previous assumption,
%2 MUL instructions scheduled in parallel benefit from amortization and cost $\alpha$, however a MUL and a MUX instructions scheduled
%in parallel still cost $2\alpha$. 
%\ishaq{We need to reword this assumption to something like "$K$ parallel MUL costs much less than $K$ (because they get amortized), but any mix of $K$  MUL and MUX still cost roughly $K$. Specifically, we should take away the low constants (2 and 1) because we know for low constants this is not true.} \ana{Again, core assumption that MUL and MUX don't benefit from amortization stays. We have to replace constant costs with functions, as in (3).}
\end{enumerate}

@@

\begin{figure*}
  \includegraphics[width=1.0\linewidth]{figs_paper_SIMD/compiler_framework.png}
  \caption{Compiler Framework. \ana{Change figure since we are not doing div-and-conquer.}}
  \label{fig:compiler_framework}
\end{figure*}

@@

@@

\figref{fig:compiler_framework} presents an overview of our compiler. This section describes the front-end phases of the compiler
and Section~\secref{sec:vectorization} describes our vectorization algorithm. \ana{We'll need a section on MOTION code gen too.}
%\ana{Need to state this more precisely.}

\subsection{Syntax and Semantic Restrictions}

Source syntax is essentially standard IMP syntax as shown below:
%\begin{figure}[t]
{\small
\[
\begin{array}{l@{~~~~~}l}
  \begin{array}{l@{~}l@{~~~}l}
 e & ::= e \; \mathit{op} \; e \mid e \mid \x \mid  \texttt{const} \mid \texttt{A[$e$]} & \mathit{expression} \\
 % s &::= t \;\f & \mathit{field} \\
 %\md &::= t\;{\m}(t \;{\this}, \; t\;\x) \; \{\; \ol{t\;\y}\;\s;\;\return\;\y \; \} 
 %                 & \mathit{method} \\ 
  s & ::= s;s \mid  & \mathit{sequence} \\
  & \x = e \mid \texttt{A[$e$]} = e \mid & \mathit{assignment \; stmt} \\
  & \texttt{for $i$ in range($I$)}: s \mid & \mathit{for \; stmt} \\
  & \texttt{if $e$ then $s$ else $s$} & \mathit{if \; stmt} \\ 
%   t &::= q \; {\sf C} & \mathit{qualified \; type}\\
%   q &::= \high \mid \poly \mid \low & \mathit{FlowCFL \; qualifier}\\
%    t' &::= q' \; \C & \mathit{field \; qualified \; type}\\
%   q' &::= \tainted \mid \poly & \mathit{field \; qualifier}\\

  \end{array}
\end{array}
\]
}
%\end{figure}

The syntax allows for array accesses, arbitrarily nested loops, and if-then-else control flow. 
Expressions are typed $q \; \tau$, where qualifier $q$ and base type $\tau$ are:

{\small
\[
\begin{array}{l@{~~~~~}l}
  \begin{array}{l@{~}l@{~~~}l}
 \tau & ::= \texttt{int}  \mid \texttt{bool} \mid \texttt{list[int]} \mid  \texttt{list[bool]} & \mathit{base \; types} \\
  q & ::= \texttt{shared} \mid \texttt{plain} & \mathit{qualifiers} \\
  \end{array}
\end{array}
\]
}

The type system is mostly standard, and in our experience, a sweet spot between simplicity and expressivity.
The \texttt{shared} qualifier denotes shared values, i.e., ones shared among the parties and computed upon 
under secure computation protocols vs. \texttt{plain} which denotes plaintext values. Shared lists denote 
shared elements, though the bounds of the list are plaintext. \ana{add subtyping}

Our compiler imposes certain semantic restrictions that it enforces throughout the various 
phased of compilation. We note that in some cases, the restrictions 
can be easily lifted and we plan to do so in future iterations of our compiler.

\begin{enumerate}
\item Loops are of the form \texttt{0 <= i < I} and bounds are fixed at compile time. 
It is a standard restriction in MPC that the bounds must be known at circuit-generation time.
\item Arrays are one-dimensional. N-dimensional arrays are linearized and accessed 
in row-major order and at this point the programmer is responsible for linearization
and access. (This restriction can be easily lifted.)
\item Array subscrpts are plaintext values. The typing rule for array access \texttt{A[$e$]} \ana{???}
Our compiler allows for output (write) arrays, however it restricts
write access to \emph{canonical writes} along the dimensions of the array. I.e., \texttt{A[i,j] = ...}
where \texttt{i} and \texttt{j} loop over the two dimensions of \texttt{A} is allowed, 
but \texttt{A[i,j+2] = ...} is not allowed. Read access can be an arbitrary function of the 
indices of the enclosing loops. 
\item \ana{Add rules for logical ops}
\item \ana{Add rules for arithmetic ops}
\item \ana{Add rules for MUX}
\end{enumerate}

We write $i,j,k$ to denote the loop nest: $i$ is the outermost loop, $j$, is immediately nested in $i$, and so on until $k$
and we use $I,J,K$ to denote the corresponding upper bounds. For simplicity, we write $A[i,j,k]$ to denote canonical access 
to an array element. In the program, canonical access is achieved via the standard row-major order formula: $(J*K)*i + K*j + k$.
To simplify the presentation we describe our algorithms in terms of three-element tuples $i,j,k$.
All discussion generalizes to arbitrarily large loop nests.

@@

\begin{enumerate}
  \item
        \textbf{Parsing}:
        Use Python's \texttt{ast} module to parse the input source code to a Python AST.
  \item
        \textbf{Syntax checking}:
        Ensure that the AST matches a restricted subset that our compiler supports.
        This step outputs an instance of the \texttt{restricted\_ast.Function} class, which represents our restricted subset of the Python AST.
  \item
        \textbf{3-address CFG conversion}:
        \benjamin{TODO: Is this a good amount of detail?} \ana{This is good amount of detail for the Tech report. In the actual paper we'll shorten a lot and add pointers to the Tech report.}

@@

Then, the control-flow graph is \emph{linearized} into MPC Source,
which has loops but no if-then-else-statements.
This means that both branches of all if-statements are executed,
and the MUX nodes determine whether to use results from the then-block or from the else-block.
The compiler linearizes the control-flow graph with a variation of breadth-first search.
Blocks with the ``merge condition'' property are only considered
the second time they are visited,
since that will be after both branches of the if-statement are visited.
Each time the compiler visits a block,
it adds the block's statements to the MPC source.
If the block ends in a for-instruction,
the compiler recursively converts the body and code after the loop to MPC source
and adds the for-loop and code after the loop to the main MPC source.
If the block does not end in a for-instruction,
the compiler recursively converts all successor branches to MPC source and
appends these to the main MPC source.

@@

An important component of our algorithm is the scalar expansion to the corresponding loop dimensionality. 
For example, \texttt{d = S[i * D + j] - C[j]} equiv. to \texttt{d = S[i,j] - C[j]}, which gave rise to \texttt{N*D} subtraction operations in the sequential schedule, 
is lifted. The argument arrays \texttt{S} and \texttt{C} are lifted and the scaler \texttt{d} is lifted: \texttt{d[i,j] = S[i,j] - C[i,j]}.
The algorithm then detects that the statement can be vectorized.
